---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1. Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

2. Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
library(ggplot2)
```

3. Make a repository in the https://github.com/stat545ubc-2021 Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Vincenzo as collaborators. 

4. When you go to submit, submit a URL to your repository to canvas. 

# Instructions

More details regarding the instructions and points allocated to each task can be found below. Briefly,

+ Each milestone is worth 30 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

+ 10 points will be allocated to the reproducibility, cleanliness, and coherence of the analysis. While the three milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as 3 chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with your dataset of choosing
+ Think of 4 questions that you would like to Answers with your data
+ Generate a reproducible and clear report using R Markdown

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

+ *apt_buildings*: Acquired courtesy of The City of Toronto’s Open Data Portal. It currently has 3455 rows and 37 columns.

+ *building_permits*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 20680 rows and 14 columns.

+ *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

+ *flow_sample*: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.

+ *parking_meters*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.

+ *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

+ *vancouver_trees*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

+ We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

+ Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data. \

**Answers for 1.1** \
1: apt_buildings    
2: steam_games    
3: vancouver_trees    
4: flow_sample \    

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*. 

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you, but do you want to use more than one? Would you like to write more comments outside of the code chunk? 

**Answers for 1.2**\
Four code chunks were used to find out the 3 attributes about each datasets.\
*apt_buildings*: includes 3455 row and 37 variables. The name and type of variables were listed below. *<dbl>* represents double-precision floating-point format, including id, no_of_elevators, rsn and so on. *<chr>* represents character format, including air_conditioning, amenities, balconies and so on.
```{r}
### EXPLORE apt_buildings ###
pillar::glimpse(apt_buildings)
```
*steam_games*: includes 40833 row and 21 variables. The name and type of variables were listed below. *<dbl>* represents double-precision floating-point format, including id, achievements, original_price, and so on. *<chr>* represents character format, including url, types, name, and so on.

```{r}
### EXPLORE steam_games ###
pillar::glimpse(steam_games)
```
*vancouver_trees*: includes 146611 row and 20 variables. The name and type of variables were listed below. *<dbl>* represents double-precision floating-point format, including tree_id, civic_number, on_street_block, and so on. *<chr>* represents character format, including std_street, genus_name, species_name, and so on. *<date>* represents date format, including date_planted.

```{r}
### EXPLORE vancouver_trees ###
pillar::glimpse(vancouver_trees)
```
*flow_sample*: includes 218 row and 7 variables. The name and type of variables were listed below. *<dbl>* represents double-precision floating-point format, including year, month, day, and flow. *<chr>* represents character format, including station_id, extreme_type, and sym.

```{r}
### EXPLORE flow_sample ###
pillar::glimpse(flow_sample)
```

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation. 

**Answers for 1.3**\
*steam_games* and *flow_sample* were chosen for further analysis. \
First, I chose *steam_games*, because this dataset provided games' reviews, which is very interesting as an indicator for the quality of a game. Besides games' reviews, games' descriptions were provided. Combining natural language processing (NLP) techniques, we can qualify descriptions and extract the topics of the games, which uncovers the development of video games. 
```{r}
head(steam_games$all_reviews)
steam_games$game_description[1]
```
Second, *flow_sample* was selected. Although this dataset is sample, it included very useful and understandable information. The flow variable can be used for provide the distribution of annual extremes. The month of extremes can be used for counting the occurrence of extremes. And two type of extremes can be treated as a filter for further analysis. 

```{r}
summary(flow_sample$flow)
table(flow_sample$month)
table(flow_sample$extreme_type)
```


1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to Answers with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

**Answers for 1.4**\

Based on the overview of summary function over the different columns, *flow_sample* was used for the following exercise. First fo all, it is easy to understand. Second, we don't need to transform the character format into numeric format, which is beyond this course. Third, we can explore several interesting and meaningful questions: (1) What is the maximum and minimum of the flow? (2) What is the range of flow in a year? (3) What is the relationship between the flow and month? All these questions can be very useful for the hydrology and tourism as this station is located on BOW RIVER AT BANFF.
```{r}
summary(flow_sample)
```


# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to Answers during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only Answers *4*. Use *dplyr* and *ggplot*.
1. Plot the distribution of a numeric variable.
2. Create a new variable based on other variables in your data (only if it makes sense)
3. Investigate how many missing values there are per variable. Can you find a way to plot this?
4. Explore the relationship between 2 variables in a plot.
5. Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6. Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7. Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8. Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to Answers with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->
________________________________________

**Answers for Task 2**\
In this question, 2.1.1, 2.1.2, 2.1.5, and 2.1.6 was chosen.

2.1.1 Plot the distribution of a numeric variable. \

```{r}
# plot a histogram
ggplot(flow_sample, aes(flow, fill=extreme_type)) +
   geom_histogram()
```

**Explanation:** The flow ($m^3/3$) was chosen for plotting the distribution. The distribution of the extreme flow is a popular research topic in hydrology. Brunner et al. (2019) mentioned that extreme low and high flows can affect economic, social, and ecological effects. Histogram was used to plot the distribution of flow here. 

Brunner, M. I., Farinotti, D., Zekollari, H., Huss, M., & Zappa, M. (2019). Future shifts in extreme flow regimes in Alpine regions. Hydrology and Earth System Sciences, 23(11), 4471-4489.



2.1.2 Create a new variable based on other variables in your data (only if it makes sense)

```{r}
# plot a create a new variable
flow_sample_new <- flow_sample %>%
  pivot_wider(id_cols = c(year,station_id), names_from = extreme_type,values_from = flow)%>%
  mutate(extreme_range= maximum-minimum)
# Note that the variable can be put back to the original dataset by pivot_longer(). But it is not required in real analysis.

# plot a histogram
ggplot(flow_sample_new, aes(extreme_range)) +
   geom_histogram()
```

**Explanation:** The range of extreme flow was also an interesting topic in hydrology. It can be used for the evaluation of the recovery of an ecosystem (Rathburn et al., 2013). Here, the station's id and year was used to change the data to wide format to create two variable: maximum and minimum. Then the range of extreme flow in a year can be calculated via the `mutate()` function. Histogram was also used to plot the distribution of the range of extreme flow here. 


Rathburn, S. L., Rubin, Z. K., & Wohl, E. E. (2013). Evaluating channel response to an extreme sedimentation event in the context of historical range of variability: Upper Colorado River, USA. Earth Surface Processes and Landforms, 38(4), 391-406.


2.1.5 Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...


```{r}
# plot a dot plot
ggplot(flow_sample, aes(x = year, y= flow, colour=extreme_type)) + 
    geom_point()

# create a new dataset for maximum
maximum_flow <- flow_sample %>%
    filter(extreme_type == "maximum")
# summary(maximum_flow$flow)
psych::describe(maximum_flow$flow)

# create a new dataset for minimum
minimum_flow <- flow_sample %>%
    filter(extreme_type == "minimum")
# summary(maximum_flow$flow)
psych::describe(minimum_flow$flow)

```

**Explanation:** As the following plot showed, the distribution maximum and minimum flows are very different. It may not make sense to put them together for further descriptive statistical analysis. Here the data was filtered by the extreme_type. Then `decribe()` can provide more statistical indices for maximum and minimum.

2.1.6 Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!


```{r}
# just for demonstration
ggplot(flow_sample, aes(extreme_type,flow)) + 
    geom_boxplot(width = 0.2)

# boxplot for maximum
ggplot(maximum_flow, aes(x = as.factor(month), y = flow)) + 
    geom_boxplot(width = 0.2)

# you can check minimum in a same way
# ggplot(minimum_flow, aes(x = as.factor(month), y = flow)) + 
#     geom_boxplot(width = 0.2)
```

**Explanation:** The boxplot can be used for checking the different observation of flow. However, the distribution of maximum and minimum are very different as mentioned before. It is not interesting to show it. Hence, the filtered dataset for maximum was used for boxplot. And this may indicated the maximums may vary by month, as shown by the median and interquartile range. ANOVA or t-test can be used for further statistical inference.   

____________________________
**Answers for Task 3**\
As what I have mention in Task 2, four research questions can be provided:\

**Question 1:** What is the distribution of the extreme flow? According to the station ID, we can find this is the flow of bow river at Banff. Understanding and visualizing the distribution of the extreme flow can the instruct local tourism and keep tourists' safety. 

**Question 2:** How did the range of the extreme flow change in a year? The range of the extreme flow can help us better understand the distribution of the extreme flow, and this also can be an index for evaluation the local ecosystem. And we can provide more information based on descriptive statistics. 

**Question 3:** What is the distribution of the maximum and minimum of the flow? The maximum and minimum of the flow may be interesting indices for the climate change, including global warming and weather extremes. The time series analysis may be useful to provide more information about this question. 


**Question 4:** Did the maximum of flow relate to month? By checking the boxplot of the maximum of flow, we found the distributions among months were different. This may indicates an interesting climatic mechanism behind the precipitation and time. ANOVA and t-test can help to provide statistical inference for this question. 






### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
